### 2022-11-24 manuscript.Rmd
Adapt rmd file to sim with 1000 iterations, fix some typos

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB is a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). AP is a fellow of the International Max Planck Research School on Computational Methods in Psychiatry and Ageing Research (COMP2PSYCH, https://www.mps-ucl-centre.mpg.de/comp2psych). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC).
  The latest supplementary materials for this article are availible at <https://github.com/ebuchberger/Estimating-Power-for-SEM>. The version of code and data used as the basis for this PDF is "`r paste0(repro::current_hash(backend = 'gert'))`".

abstract: |
  Determining the compositional structure and dimensionality of psychological constructs lies at the heart of many research questions in developmental science. Structural equation modelling (SEM) provides a versatile framework for formalizing and estimating the relationships among multiple latent constructs. While the flexibility of SEM can accommodate many complex assumptions on the underlying structure of psychological constructs, it makes a-priori estimation of statistical power and required sample size challenging. This difficulty is magnified when comparing non-nested SEMs, which prevents the use of the likelihood-ratio test. Sample size estimates for SEM model fit comparisons typically rely on generic rules of thumb. Such heuristics can be misleading because statistical power in SEM depends on a variety of model properties. Here, we demonstrate a Monte Carlo simulation approach for estimating a-priori statistical power for non-nested model comparisons in an SEM framework. We provide a step-by-step guide to this approach based on an example from our memory development research in children.

@@

Since Cohen’s classic study on statistical power [@cohen_1962], numerous researchers from psychological and neurocognitive science have stressed the issue of insufficient statistical power and its consequences on the interpretation of scientific findings in the face in inadequate sample sizes [@anderson_2017; @button_2013; @sedlmeier_1989, @vankov_2014]. However, determining required sample size in an SEM approach is far from trivial. While the investigation of target effects in SEM can be achieved via analytical computations [@satorra_1985], tackling more complex research questions that rely on the comparison of non-nested models poses a challenge in determining sufficient sample sizes.

In this study, we implemented a Monte Carlo based simulation approach and provide a step-by-step guide for conducting randomization-based analyses for a-priori sample size estimations for non-nested model comparison. Specifically, we exemplify this approach with a specific research question, focusing on the componential structure of memory processes in early childhood [for theoretical background see @buchberger_2022]. Memory developmental research represents an excellent example for this methodological approach, as researchers from different fields have put forth competing ideas on the underlying structure of memory in childhood: While a rich body of empirical work from many decades has focused on the dichotomy between episodic and semantic memory [@tulving_1972], more recent neurocomputational theories have introduced a process-focused approach to memory development. Tenets from contemporary memory models invite the hypothesis that three kinds of neurocomputations may complementarily support adaptive memory functioning, thereby assuming a tri-partite structure of memory [@mcclelland_1995;@norman_2003; @schapiro_2017]. Such competing theories on the underlying structure of memory can be addressed via model comparisons in an SEM framework. In this tutorial, we address two overarching issues that affect estimates of statistical power for such comparisons: (i) the separability of the theoretical constructs and (ii) the reliability with which the constructs are being measured. We employ this example to demonstrate the procedure and the utility of such methods, with the aim of illustrating a transferable approach to other research questions in different domains and disciplines.

@@

The relationship among manifest and latent variables can be described in a mathematical model, which specifies all assumed parameters. Two models M1 and M2 are referred to as nested if the parameter space in a more restrictive model M2 represents a subspace of the parameter space of the more general model M1  [@bentler_1980]. This means that the two models only differ with regards to the specification of one or multiple parameters. Nesting of models is usually achieved via constraining free parameters from M1 to equality or to known constants. The concept of nesting represents a convenient characteristic when comparing competing models, as it allows the use of a Likelihood Ratio test to evaluate their relative model fits [@satorra_1985]. When comparing non-nested models, researchers usually have to revert to model fit indices that indicate how well a given model describes the data, such as the Bayesian Information Criterion [BIC, @raftery_1986; @schwarz_1978], the Akaike Information Criterion [AIC, @akaike_1974], the Root Mean Square Error of Approximation [RMSEA, @steiger_2016], and the Comparative Fit Index [CFI, @bentler_1990]. Such indices can guide heuristic decisions that do not hold any statistical guarantees (e.g., a pre-specified Type-I error level).

## 1.2	Statistical Power in SEM
Over the past decades, the issue of low statistical power and its consequences for the interpretation of scientific findings has gained awareness in neurocognitive and psychological research [@cohen_1988; @maccallum_2000; @maxwell_2004; @rossi_1990]. Statistical power refers to the probability of rejecting a null hypothesis, when it is indeed false (i.e. the probability of not committing a Type-II error) and is directly linked to the sample size of a study, the magnitude of the targeted effect, and the reliability of measurement [@brandmaier_2018; @cohen_1988]. While low statistical power decreases the scientific utility of any given study, it also decreases the likelihood that a given significant result actually reflects a true effect [@button_2013]. To allow well designed empirical research to detect effects in the sample under investigation, a-priori estimates of statistical power (that is computation of power estimates before conducting the study) and thus informed decisions on required sample size are crucial to prevent underpowered research [@button_2013]. While the desired level of statistical power can depend on specific aspects of the research question at hand, a typical convention for adequate power in the behavioral sciences is 0.8 [@cohen_1988]. However, others have argued that there is no reason to prefer Type-I errors over Type-II errors and thus one should better aim for a statistical power of 0.95, if the level of significance is kept at the conventional 5%. Note that of course, that considerations on statistical power and thus sample size estimates need to be balanced with the probability of committing Type-I errors, that is the probability to reject a null hypothesis if it is indeed true.

@@

### 1.2.1	Statistical Power to Detect a Specific Effect of Interest
Central to many studies that employ SEM is the question of whether a specific parameter in a model (e.g., strength of a specific correlation or the slope parameter in a latent regression model) is different from a given value [e.g. @canada_2021]. Imagine that a researcher aims to test whether the correlation between two latent constructs significantly differs from zero. To this end, they would compare the model fit between one model in which the correlation parameter is freely estimated and another model in which it is fixed to zero [@satorra_1985]. The corresponding null hypothesis in this case states that the parameter restrictions hold in the population. The difference in model fit will follow a $\chi^2$-distribution with degrees-of-freedom (*df*) equal to the difference of freely estimated parameters between the two models if the null hypothesis is true [@neale_2000, in this example *df* = 1, because only the correlation between the two latent factors is fixed in the restricted model]. If restricting the parameter of interest results in a significantly poorer model fit, this would suggest that the respective parameter indeed significantly differs from zero.
Following the logic outlined above, researchers can determine the statistical power for detecting a target effect in an SEM framework before conducting the study. A key aspect for this a-priori power calculation hinges on translating the differences in a specific parameter estimate into an effect size. Such translation can be achieved by investigating the discrepancy between the model-implied variance-covariance matrices associated with (1) the population model including the “true” parameter values and (2) the hypothesized model. The discrepancy between both matrices is quantified based on a specific fit function [for details on fit functions see @bollen_1989]. Recently, several user friendly tools have emerged that can guide modelers in quantifying effect sizes of a target effect to estimate statistical power in SEM [e.g. the R packages semPower @jobst_2021; or power4SEM @jak_2020 for analytical approaches; and the Shiny app pwrSEM @wang_2021; the interactive study planner tool LIFESPAN @brandmaier_2015; or the R package simsem  @pornprasertmanit_2021 for simulation-based approaches]. As the investigation of a target effect practically translates into the comparison of nested models, researchers can in these cases analytically determine the statistical power and use this information for decisions on required sample size [@satorra_1985].

@@

An alternative prominent view is the classic dichotomy between episodic and semantic memory systems [@squire_1987; @tulving_1972]. Such a bi-partite architecture of memory postulates a division between one component responsible for learning specific events embedded in their temporal and spatial context (episodic memory) and a second component responsible for learning regularities and extracting generalized knowledge (semantic memory). Previous research comparing a unitary vs. a bi-partite account of memory in an SEM framework has supported the idea of separable memory factors underlying declarative memory, at least in adults [@nyberg_1994]. 
Adopting a process-oriented view, recent computational models of memory have argued for a functionalist distinction between memory specificity and generalization through a labor division between the hippocampus and the cortex [@mcclelland_1995; @norman_2003]. Here, a set of neurocomputations support different memory demands. Important to learning specific episodes are pattern separation that discriminates between similar memories through the reduction of representational similarity, and pattern completion that retrieves linked associations among co-occurring elements [@marr_1971; @norman_2003; @rolls_2016]. Both of these computations are thought to be specialties of the hippocampus. In contrast, the cortex is well suited to slowly learn statistical regularities that enables generalization [@mcclelland_1995; @norman_2003]. Interestingly, rapid generalization also relies on the hippocampus, either via retrieval mechanisms of related episodes (Kumaran & McClelland, 2012) or via a distributed coding scheme carried by a specific subset of the hippocampal circuitry [@schapiro_2017]. From this vantage point, memory abilities take shape of a tri-partite structure, encompassing pattern separation, pattern completion and generalization as three separable mnemonic processes.

@@

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (section 3.1) and the reliability of the manifest variables (section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different iterations of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (section 3.3).

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [PS]; Pattern Completion, [PC]; Generalization, [GEN]) which are measured through the manifest variables (Tasks 1 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

@@

### 2.1.2 Step 2: Simulation Design
When defining the structure of the theoretical model for data generation, it is crucial to specify all parameters in the model, particularly the loading strength of the manifest variables, along with the means and (co-)variances of the latent factors. This step also requires the specification of which parameter values should be kept constant or modified across different simulation iterations. Parameters for which there are strong assumptions (e.g. derived from previous work) can be entered in the models and kept constant. Parameters for which the impact on statistical power is of interest should be included as a condition in the simulation design. For example, if the question pertains to how the reliability of the measures may impact the estimates of required sample size, different levels of loading strength can be included in the conditions for which data is being simulated. The createDesign() function of the simDesign package can be used to create a full list of conditions for the simulation. All parameters that should be varied in the simulation are included as arguments in the createDesign() function. Full crossings of all possible values for the different parameters are then generated as conditions for the simulation. Note, however, that the number of simulation conditions and therefore the required computation time increases exponentially for full combinations of all simulation conditions. Researchers should thus be mindful about the number of conditions in a given simulation to avoid overly computationally intensive simulations, which further can be increasingly difficult to interpret. If only specific combinations of the parameters in the simulation are of interest for a given research question, the resulting design data frame can be adjusted accordingly, e.g., by deleting specific rows from the data frame. Below is our design data frame that includes different simulation conditions for (i) the loading strength of the manifest variables and (ii) the sample sizes:

@@

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as a for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as b to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). The population model is defined by a model string, which specifies three latent factors (ps, pc, gen), measured by the 13 manifest indicators (task1 – task13). The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The covariance between the latent factors is set to the respective value from the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package and stored as dat.

@@

In a final step, we call the runSimulation() function from the simDesign package to execute the consecutive steps outlined above. Running the simulation for our example could look like this:
```{r run_simulation_cov_model3, eval = FALSE, echo = TRUE}
```
In this function, the pre-defined functions generate_data(), analyze_results() and summarise_results() are called. The number of replications is specified in the replications argument. To ensure a reasonable stability of the results, a sufficient number of replications is recommended. In our example, we simulated 10,000 datasets per condition. This leads to a standard error of the simulated power of less than 0.5%. The store_results argument can be used to store the individual results generated by the summarize function, allowing users to investigate the outcome on every iteration (in this example, the data frame containing all fit measures and convergence information for each iteration). Note however, that storing the individual results increases computation time and size of the resulting outcome arguments. The argument parallel can be used for parallel processing across multiple cores (e.g. when running the analysis on a computer cluster) to reduce computation time. If not specified otherwise, all available cores will be used per default when parallel = TRUE.

@@

In the example of comparing competing models of memory structure, we first examined the separability of the theoretical models by investigating the influence of the latent factors’ correlations on statistical power estimates. In our example simulation, we investigated the statistical power to retrieve Model 3 as the best fitting model, when it indeed generated the data, while varying sample sizes from 50 to 200 and factor inter-correlations from 0.1 to 0.9. In this case, we restricted the simulated correlations between the latent factors to positive values as there is no indication in the literature to suspect a negative relation among these memory capacities [e.g., @tucker-drob_2019]. Figure 3A displays the obtained power curves illustrating the estimated statistical power dependent on the total sample size and the factor inter-correlation. For this estimate, the loading of the manifest variables was fixed at 0.7 and the inter-correlation between all latent factors were varied simultaneously.  
As expected, statistical power increased with increasing sample size. However, the separability of the latent factors mattered: The higher the correlations between factors, the less able we were to distinguish multi-factor models from a one-factor model. That is, the smaller the difference between competing models, the greater the sample size that would be required to dissociate between them with adequate power. For very high levels of inter-correlation (0.8 and higher) the increase of statistical power with increasing sample size was remarkably lower than for lower levels of inter-correlation (0.7 and below). When the inter-factor correlation was 0.9, not even a larger sample size (*N* =200) achieved an adequate level of power of 80%. These results are not surprising given that increasing the inter-correlations between the latent factors renders the models more similar to each other, and thus makes their distinction more challenging. For moderate levels of inter-correlation (<= 0.7) a statistical power of 95% was reached for *N* = 120.

@@

In a second step, we further examined the effect of the construct reliability on statistical power by investigating the loading strength of the manifest variables. To this end we conducted a second simulation following the steps outlined above while varying the sample size from 50 to 200 and the loadings of the manifest variables from 0.5 to 0.9. Figure 3B displays the obtained power curves, illustrating statistical power dependent on total sample size and loading strength of the manifest variables. For this estimate, the factor inter-correlation was set to 0.3. Again, increasing sample size lead to higher statistical power. However, the loading strength impacted estimates of statistical power: the higher the loadings of the manifest variables, the higher the statistical power achieved by a given sample size. For reasonably high levels of factor loadings (>= 0.7) a statistical power of 95% was reached for *N* = 120.

@@

```{r, echo = FALSE, fig.width=11.3, fig.height=4.3, fig.align = 'center', fig.cap = "Simulation results: Estimated statistical power for recovering the data-generating model, i.e. Model 3 (y-axis) as a function of sample size (x-axis) with varying the factor inter-correlations when the loading strength was fixed to 0.7 (A) $;$ and with varying the loading strength when the factor inter-correlation was fixed to 0.3 (B). For each combination of parameters, 10,000 datasets were simulated. The red dashed horizontal line positioned at 95\\% discovery rate reflects the cut-off for identifying the appropriate sample size. The syntax to generate these plots with ggplot() can be found in the supplementary R Code on GitHub"}

@@

The results of this analysis reveal that even for low levels of loadings strength (< 0.7), the probability of erroneously picking Model 3 as the best fitting model is reasonably low (< 7%). Likewise, this probability remains reasonably low for different levels of inter-factor correlations for all sample sizes (< 5%). Thus, in this specific model comparison Type-I error probability does not appear to be the main aspect driving sample size requirements for an informative research design.

>>>>

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB is a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). AP is a fellow of the International Max Planck Research School on Computational Methods in Psychiatry and Ageing Research (COMP2PSYCH, https://www.mps-ucl-centre.mpg.de/comp2psych). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC).
  The latest supplementary materials for this article are available at <https://github.com/ebuchberger/Estimating-Power-for-SEM>. The version of code and data used as the basis for this PDF is "`r paste0(repro::current_hash(backend = 'gert'))`".

abstract: |
  Determining the compositional structure and dimensionality of psychological constructs lies at the heart of many research questions in developmental science. Structural equation modelling (SEM) provides a versatile framework for formalizing and estimating the relationships among multiple latent constructs. While the flexibility of SEM can accommodate many complex assumptions on the underlying structure of psychological constructs, it makes a-priori estimation of statistical power and required sample size challenging. This difficulty is magnified when comparing non-nested SEMs, which prevents the use of the Likelihood Ratio test. Sample size estimates for SEM model fit comparisons typically rely on generic rules of thumb. Such heuristics can be misleading because statistical power in SEM depends on a variety of model properties. Here, we demonstrate a Monte Carlo simulation approach for estimating a-priori statistical power for non-nested model comparisons in an SEM framework. We provide a step-by-step guide to this approach based on an example from our memory development research in children.

@@

Since Cohen’s classic study on statistical power [@cohen_1962], numerous researchers from psychological and neurocognitive science have stressed the issue of insufficient statistical power and its consequences on the interpretation of scientific findings in the face in inadequate sample sizes [@anderson_2017; @button_2013; @sedlmeier_1989; @vankov_2014]. However, determining required sample size in an SEM approach is far from trivial. While the investigation of target effects in SEM can be achieved via analytical computations [@satorra_1985], tackling more complex research questions that rely on the comparison of non-nested models poses a challenge in determining sufficient sample sizes.   
In this study, we implemented a Monte Carlo based simulation approach and provide a step-by-step guide for conducting randomization-based analyses for a-priori sample size estimations for non-nested model comparison. Specifically, we exemplify this approach with a specific research question, focusing on the componential structure of memory processes in early childhood [for theoretical background see @buchberger_2022]. Memory developmental research represents an excellent example for this methodological approach, as researchers from different fields have put forth competing ideas on the underlying structure of memory in childhood: While a rich body of empirical work from many decades has focused on the dichotomy between episodic and semantic memory [@tulving_1972], more recent neurocomputational theories have introduced a process-focused approach to memory development. Tenets from contemporary memory models invite the hypothesis that three kinds of neurocomputations may complementarily support adaptive memory functioning, thereby assuming a tri-partite structure of memory [@mcclelland_1995;@norman_2003; @schapiro_2017]. Such competing theories on the underlying structure of memory can be addressed via model comparisons in an SEM framework. In this tutorial, we address two overarching issues that affect estimates of statistical power for such comparisons: (i) the separability of the theoretical constructs and (ii) the reliability with which the constructs are being measured. We employ this example to demonstrate the procedure and the utility of such methods, with the aim of illustrating a transferable approach to other research questions in different domains and disciplines.

@@

The relationship among manifest and latent variables can be described in a mathematical model, which specifies all assumed parameters. Two models M1 and M2 are referred to as nested, if the parameter space in a more restrictive model M2 represents a subspace of the parameter space of the more general model M1  [@bentler_1980]. This means that the two models only differ with regards to the specification of one or multiple parameters. Nesting of models is usually achieved via constraining free parameters from M1 to equality or to known constants. The concept of nesting represents a convenient characteristic when comparing competing models, as it allows the use of a Likelihood Ratio test to evaluate their relative model fits [@satorra_1985]. When comparing non-nested models, researchers usually have to revert to model fit indices that indicate how well a given model describes the data, such as the Bayesian Information Criterion [BIC, @raftery_1986; @schwarz_1978], the Akaike Information Criterion [AIC, @akaike_1974], the Root Mean Square Error of Approximation [RMSEA, @steiger_2016], and the Comparative Fit Index [CFI, @bentler_1990]. Such indices can guide heuristic decisions that do not hold any statistical guarantees (e.g., a pre-specified Type-I error level).

## 1.2	Statistical Power in SEM
Over the past decades, the issue of low statistical power and its consequences for the interpretation of scientific findings has gained awareness in neurocognitive and psychological research [@cohen_1988; @maccallum_2000; @maxwell_2004; @rossi_1990]. Statistical power refers to the probability of rejecting a null hypothesis, when it is indeed false (i.e. the probability of not committing a Type-II error) and is directly linked to the sample size of a study, the magnitude of the targeted effect, and the reliability of measurement [@brandmaier_2018; @cohen_1988]. While low statistical power decreases the scientific utility of any given study, it also decreases the likelihood that a given significant result actually reflects a true effect [@button_2013]. To allow well designed empirical research to detect effects in the sample under investigation, a-priori estimates of statistical power (that is computation of power estimates before conducting the study) and thus informed decisions on required sample size are crucial to prevent underpowered research [@button_2013]. While the desired level of statistical power can depend on specific aspects of the research question at hand, a typical convention for adequate power in the behavioral sciences is 0.8 [@cohen_1988]. However, others have argued that there is no reason to prefer Type-I errors over Type-II errors and thus one should better aim for a statistical power of 0.95, if the level of significance is kept at the conventional 5%. Note that of course, considerations on statistical power and thus sample size estimates need to be balanced with the probability of committing Type-I errors, that is the probability to reject a null hypothesis if it is indeed true.

@@

### 1.2.1	Statistical Power to Detect a Specific Effect of Interest
Central to many studies that employ SEM is the question of whether a specific parameter in a model (e.g., strength of a specific correlation or the slope parameter in a latent regression model) is different from a given value [e.g. @canada_2021]. Imagine that a researcher aims to test whether the correlation between two latent constructs significantly differs from zero. To this end, they would compare the model fit between one model in which the correlation parameter is freely estimated and another model in which it is fixed to zero [@satorra_1985]. The corresponding null hypothesis in this case states that the parameter restrictions hold in the population. The difference in model fit will follow a $\chi^2$-distribution with degrees-of-freedom (*df*) equal to the difference of freely estimated parameters between the two models, if the null hypothesis is true [@neale_2000, in this example *df* = 1, because only the correlation between the two latent factors is fixed in the restricted model]. If restricting the parameter of interest results in a significantly poorer model fit, this would suggest that the respective parameter indeed significantly differs from zero.
Following the logic outlined above, researchers can determine the statistical power for detecting a target effect in an SEM framework before conducting the study. A key aspect for this a-priori power calculation hinges on translating the differences in a specific parameter estimate into an effect size. Such translation can be achieved by investigating the discrepancy between the model-implied variance-covariance matrices associated with (1) the population model including the “true” parameter values and (2) the hypothesized model. The discrepancy between both matrices is quantified based on a specific fit function [for details on fit functions see @bollen_1989]. Recently, several user friendly tools have emerged that can guide modelers in quantifying effect sizes of a target effect to estimate statistical power in SEM [e.g. the R packages semPower @jobst_2021; or power4SEM @jak_2020 for analytical approaches; and the Shiny app pwrSEM @wang_2021; the interactive study planner tool LIFESPAN @brandmaier_2015; or the R package simsem  @pornprasertmanit_2021 for simulation-based approaches]. As the investigation of a target effect practically translates into the comparison of nested models, researchers can in these cases analytically determine the statistical power and use this information for decisions on required sample size [@satorra_1985].

@@

An alternative prominent view is the classic dichotomy between episodic and semantic memory systems [@squire_1987; @tulving_1972]. Such a bi-partite architecture of memory postulates a division between one component responsible for learning specific events embedded in their temporal and spatial context (episodic memory) and a second component responsible for learning regularities and extracting generalized knowledge (semantic memory). Previous research comparing a unitary vs. a bi-partite account of memory in an SEM framework has supported the idea of separable memory factors underlying declarative memory, at least in adults [@nyberg_1994].

Adopting a process-oriented view, recent computational models of memory have argued for a functionalist distinction between memory specificity and generalization through a labor division between the hippocampus and the cortex [@mcclelland_1995; @norman_2003]. Here, a set of neurocomputations support different memory demands. Important to learning specific episodes are pattern separation that discriminates between similar memories through the reduction of representational similarity, and pattern completion that retrieves linked associations among co-occurring elements [@marr_1971; @norman_2003; @rolls_2016]. Both of these computations are thought to be specialties of the hippocampus. In contrast, the cortex is well suited to slowly learn statistical regularities that enables generalization [@mcclelland_1995; @norman_2003]. Interestingly, rapid generalization also relies on the hippocampus, either via retrieval mechanisms of related episodes (Kumaran & McClelland, 2012) or via a distributed coding scheme carried by a specific subset of the hippocampal circuitry [@schapiro_2017]. From this vantage point, memory abilities take shape of a tri-partite structure, encompassing pattern separation, pattern completion and generalization as three separable mnemonic processes.

@@

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (section 3.1) and the reliability of the manifest variables (section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different conditions of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (section 3.3).

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [ps]; Pattern Completion, [pc]; Generalization, [gen]) which are measured through the manifest variables (asks 1 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

@@

### 2.1.2 Step 2: Simulation Design
When defining the structure of the theoretical model for data generation, it is crucial to specify all parameters in the model, particularly the loading strength of the manifest variables, along with the means and (co-)variances of the latent factors. This step also requires the specification of which parameter values should be kept constant or which should be modified across different simulation iterations. Parameters for which there are strong assumptions (e.g. derived from previous work) can be entered in the models and kept constant. Parameters for which the impact on statistical power is of interest should be included as a condition in the simulation design. For example, if the question pertains to how the reliability of the measures may impact the estimates of required sample size, different levels of loading strength can be included in the conditions for which data is being simulated. The createDesign() function of the simDesign package can be used to create a full list of conditions for the simulation. All parameters that should be varied in the simulation are included as arguments in the createDesign() function. Full crossings of all possible values for the different parameters are then generated as conditions for the simulation. Note, however, that the number of simulation conditions and therefore the required computation time increases exponentially for full combinations of all simulation conditions. Researchers should thus be mindful about the number of conditions in a given simulation to avoid overly computationally intensive simulations, which further can be increasingly difficult to interpret. If only specific combinations of the parameters in the simulation are of interest for a given research question, the resulting design data frame can be adjusted accordingly, e.g., by deleting specific rows from the data frame. Below is our design data frame that includes different simulation conditions for (i) the loading strength of the manifest variables and (ii) the sample sizes:

@@

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as "a" for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as "b" to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). The population model is defined by a model string, which specifies three latent factors (ps, pc, gen), measured by the 13 manifest indicators (task1 – task13). The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The covariance between the latent factors is set to the respective value from the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package and stored as dat.

@@

In a final step, we call the runSimulation() function from the simDesign package to execute the consecutive steps outlined above. Running the simulation for our example could look like this:
```{r run_simulation_cov_model3, eval = FALSE, echo = TRUE}
```
In this function, the pre-defined functions generate_data(), analyze_results() and summarise_results() are called. The number of replications is specified in the replications argument. To ensure a reasonable stability of the results, a sufficient number of replications is recommended. In our example, we simulated 1,000 datasets per condition. This leads to a standard error of the simulated power of less than 0.5%. The store_results argument can be used to store the individual results generated by the summarize function, allowing users to investigate the outcome on every iteration (in this example, the data frame containing all fit measures and convergence information for each iteration). Note however, that storing the individual results increases computation time and size of the resulting outcome arguments. The argument parallel can be used for parallel processing across multiple cores (e.g. when running the analysis on a computer cluster) to reduce computation time. If not specified otherwise, all available cores will be used per default when parallel = TRUE.

@@

In the example of comparing competing models of memory structure, we first examined the separability of the theoretical models by investigating the influence of the latent factors’ correlations on statistical power estimates. In our example simulation, we investigated the statistical power to retrieve Model 3 as the best fitting model, when it indeed generated the data, while varying sample sizes from 50 to 200 and factor inter-correlations from 0.1 to 0.9. In this case, we restricted the simulated correlations between the latent factors to positive values, as there is no indication in the literature to suspect a negative relation among these memory capacities [e.g., @tucker-drob_2019]. Figure 3A displays the obtained power curves illustrating the estimated statistical power dependent on the total sample size and the factor inter-correlation. For this estimate, the loading of the manifest variables was fixed at 0.7 and the inter-correlation between all latent factors were varied simultaneously.  
As expected, statistical power increased with increasing sample size. However, the separability of the latent factors mattered: The higher the correlations between factors, the less able we were to distinguish multi-factor models from a one-factor model. That is, the smaller the difference between competing models, the greater the sample size that would be required to dissociate between them with adequate power. For very high levels of inter-correlation (0.8 and higher) the increase of statistical power with increasing sample size was remarkably lower than for lower levels of inter-correlation (0.7 and below). When the inter-factor correlation was 0.9, not even a larger sample size (*N* =200) achieved an adequate level of power of 80%. These results are not surprising given that increasing the inter-correlations between the latent factors renders the models more similar to each other, and thus makes their distinction more challenging. For moderate levels of inter-correlation (<= 0.7) a statistical power of 95% was reached for *N* = 125.

@@

In a second step, we further examined the effect of the construct reliability on statistical power by investigating the loading strength of the manifest variables. To this end we conducted a second simulation following the steps outlined above while varying the sample size from 50 to 200 and the loadings of the manifest variables from 0.5 to 0.9. Figure 3B displays the obtained power curves, illustrating statistical power dependent on total sample size and loading strength of the manifest variables. For this estimate, the factor inter-correlation was set to 0.3. Again, increasing sample size lead to higher statistical power. However, the loading strength impacted estimates of statistical power: the higher the loadings of the manifest variables, the higher the statistical power achieved by a given sample size. For reasonably high levels of factor loadings (>= 0.7) a statistical power of 95% was reached for *N* = 125.

@@

```{r, echo = FALSE, fig.width=11.3, fig.height=4.3, fig.align = 'center', fig.cap = "Simulation results: Estimated statistical power for recovering the data-generating model, i.e. Model 3 (y-axis) as a function of sample size (x-axis) with varying the factor inter-correlations when the loading strength was fixed to 0.7 (A) $;$ and with varying the loading strength when the factor inter-correlation was fixed to 0.3 (B). For each combination of parameters, 1,000 datasets were simulated. The red dashed horizontal line positioned at 95\\% discovery rate reflects the cut-off for identifying the appropriate sample size. The syntax to generate these plots with ggplot() can be found in the supplementary R Code on GitHub"}

@@

The results of this analysis reveal that even for low levels of loadings strength (< 0.7), the probability of erroneously picking Model 3 as the best fitting model is reasonably low (< 8%). Likewise, this probability remains reasonably low for different levels of inter-factor correlations for all sample sizes (< 5%). Thus, in this specific model comparison Type-I error probability does not appear to be the main aspect driving sample size requirements for an informative research design.

## 3.4	Deriving a Sample Estimate From the Simulation Results
To derive a joint sample size estimate from the different parameters in the simulation, it is crucial to reiterate the theoretical assumptions on the strength of all parameters investigated in the simulation. In our illustration, it is reasonable to assume that the latent factors in the tri-partite model (Model 3, Figure 1) are only moderately correlated (below 0.7). Therefore, our model comparison requires a sample size of *N* = 125 to reach statistical power of 95% to adjudicate between the theoretical models, if the hypothesized model (Model 3) generates the data. We further assume that the manifest variables capture the latent constructs reasonably well — that is, exerting a loading strength of 0.7 or above. The results of the simulation suggest that, again, *N* = 125 would suffice to achieve 95% with regard to the reliability of the manifest factors. In our example, the results of both analyses align, such that a sample size of *N* = 125 suffices to achieve a power of 95% regarding both the separability of the theoretical constructs and the reliability of the indicators. In cases where the derived sample estimate differs between the two analyses we suggest to derive the more conservative estimate of required sample size from the two simulations.
The results from the Type-I error analysis further indicate that the probability of erroneously choosing an overly complex model is generally low. Therefore, the probability of committing a Type-I error for the sample size derived from the simulations on model separability and measurement reliability (*N* = 125) is negligible.
